---
title: "RMB Package: Complete Guide"
subtitle: "R Multi-Beam Drosophila Activity Analysis"
author: "Refactored by Catbot for Yongjun Li"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    highlight: tango
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)

# Set working directory and load packages
setwd("/home/liy27/projects/RMB")

suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
})

# Source all RMB functions
source("R/constants.R")
source("R/MonitorClass.R")
source("R/transforms.R")
source("R/plotting.R")
source("R/themes.R")
source("R/plotting_advanced.R")
source("R/checkpoints.R")
source("R/pipeline.R")
source("R/utils.R")
```

\newpage

# Introduction

## What I Did: Package Refactoring

I converted the RMB project from a collection of scripts into a **proper R package** with:

### Code Consolidation
- **Before**: 12+ scattered R files with duplicate functions
- **After**: 9 organized R files with clear responsibilities

### Key Improvements Made

| Issue | Solution |
|-------|----------|
| Duplicate functions in multiple files | Consolidated into single source files |
| Magic numbers (1440, 15, 32) everywhere | Created `constants.R` with named constants |
| Fake S4 class with custom `@` operator | Created proper S4 class `MonitorS4` |
| Hardcoded file paths | Auto-detection with configurable overrides |
| No tests | Added testthat test infrastructure |
| No CI/CD | Added GitHub Actions workflows |

### New Features Added

1. **One-command pipeline**: `runPipeline()` does everything
2. **Checkpoints**: Save analysis state at each step
3. **Publication themes**: `theme_rmb()` for beautiful plots
4. **Advanced visualizations**: Actograms, sleep profiles, group comparisons

\newpage

# Package Structure

```
RMB/
├── DESCRIPTION          # Package metadata
├── NAMESPACE            # Exported functions  
├── LICENSE              # MIT license
├── R/
│   ├── constants.R      # Named constants (no magic numbers)
│   ├── MonitorClass.R   # S4 class definition
│   ├── transforms.R     # Data transformation functions
│   ├── plotting.R       # Basic plotting functions
│   ├── plotting_advanced.R  # Advanced visualizations
│   ├── themes.R         # ggplot2 themes and color palettes
│   ├── checkpoints.R    # Analysis checkpoints and summaries
│   ├── pipeline.R       # Complete analysis pipeline
│   └── utils.R          # Helper functions
├── tests/testthat/      # Unit tests
├── vignettes/           # Tutorials
└── inst/extdata/        # Example data
```

\newpage

# Step-by-Step Analysis Walkthrough

Let me walk you through the complete analysis using your test data.

## Step 1: Load Monitor Files

First, we find and load the raw DAM monitor data files.

```{r load-data}
data_dir <- "./test"

# Find monitor files
monitor_files <- loadMonitorFiles(data_dir)
cat("Found", length(monitor_files), "monitor files:\n")
for (f in monitor_files) {
  cat(" -", basename(f), "\n")
}
```

## Step 2: Create MonitorS4 Objects

Each monitor file is loaded into a `MonitorS4` object - similar to how Seurat works for single-cell data.

```{r create-monitors}
# Load first monitor as example
monitor1 <- createMonitor(monitor_files[1], verbose = TRUE)
```

The MonitorS4 object has these slots:

- `@assays`: List of data matrices (mt, pn, ct, awake, sleep)
- `@meta.data`: Fly metadata (phenotype, treatment, etc.)
- `@time`: Time information
- `@monitor_name`: Identifier

## Step 3: Process Raw Data

The `processMonitor()` function parses the raw DAM format and extracts:

- **MT (Movement)**: Beam crosses per minute
- **Pn (Position)**: Position in tube (1-15 for multi-beam)
- **CT (Cumulative)**: Cumulative activity

```{r process-monitor}
monitor1 <- processMonitor(monitor1, verbose = TRUE)
```

## Step 4: Load All Monitors and Merge

```{r merge-monitors}
# Load and process all monitors
monitor_list <- lapply(monitor_files, function(f) {
  m <- createMonitor(f, verbose = FALSE)
  m <- processMonitor(m, verbose = FALSE)
  return(m)
})

# Merge into single object
monitor <- mergeMonitors(monitor_list, verbose = TRUE)
```

\newpage

## Step 5: Quality Control - Remove Dead Flies

Flies with very low total activity are likely dead and should be removed.

```{r qc}
n_before <- ncol(monitor@assays$mt)
monitor <- removeDeadFlies(monitor, min_movement = 10, verbose = TRUE)
n_after <- ncol(monitor@assays$mt)

cat("\nRemoved", n_before - n_after, "dead flies\n")
cat("Remaining:", n_after, "alive flies\n")
```

## Step 6: Check Data Quality

```{r quality-check}
quality <- checkQuality(monitor, verbose = TRUE)
```

\newpage

## Step 7: Sleep Analysis

Drosophila sleep is defined as **5 or more consecutive minutes without any beam crosses**.

```{r sleep-analysis}
# Convert activity to awake/sleep binary
monitor@assays$awake <- activityToAwake(monitor@assays$mt, threshold = 5)
monitor@assays$sleep <- awakeToSleep(monitor@assays$awake)

# Calculate sleep statistics
total_sleep <- colSums(monitor@assays$sleep, na.rm = TRUE)
percent_sleep <- total_sleep / nrow(monitor@assays$sleep) * 100

cat("Sleep Statistics:\n")
cat("  Mean total sleep:", round(mean(total_sleep), 1), "minutes\n")
cat("  Mean % time asleep:", round(mean(percent_sleep), 1), "%\n")
cat("  Range:", round(min(percent_sleep), 1), "-", round(max(percent_sleep), 1), "%\n")
```

## Step 8: Position Analysis (Multi-Beam Only)

For multi-beam monitors, we can track where flies spend their time.

```{r position-analysis}
# Position while awake only (exclude sleeping positions)
monitor@assays$pn_awake <- maskByAwake(monitor@assays$pn, monitor@assays$awake)

# Calculate position frequency
pos_freq <- positionFrequency(monitor@assays$pn_awake)
cat("Position frequency matrix:", nrow(pos_freq), "positions x", ncol(pos_freq), "flies\n")

# Show average time at each position
avg_freq <- rowMeans(pos_freq, na.rm = TRUE)
names(avg_freq) <- paste0("Pos", 1:15)
print(round(avg_freq, 1))
```

\newpage

# Visualizations

## Activity Raster Plot

Shows activity patterns across multiple days. Each row is one day, columns are time of day.

```{r activity-raster, fig.cap="Activity raster plot showing daily activity patterns for a single fly. Darker colors indicate higher activity. Gray shading shows dark period (ZT12-24)."}
plotActivityRaster(monitor, channel = 1, 
                   title = "Activity Raster - Fly 1",
                   color_high = "#2E86AB")
```

\newpage

## Sleep Profile

Shows average sleep across the day, grouped by phenotype.

```{r sleep-profile, fig.cap="Sleep profile showing percentage of time asleep across the 24-hour day. Gray shading indicates dark period."}
# Add fake phenotype data for demonstration
n_ch <- ncol(monitor@assays$mt)
monitor@meta.data <- data.frame(
  Phenotype = rep(c("Control", "Mutant"), length.out = n_ch)
)

plotSleepProfile(monitor, 
                 bin_minutes = 30,
                 title = "Sleep Profile by Genotype")
```

\newpage

## Group Comparison - Total Activity

```{r activity-comparison, fig.cap="Total activity comparison between groups. Bars show mean ± SEM, points show individual flies."}
plotGroupComparison(monitor, 
                    metric = "total_activity",
                    title = "Total Activity by Genotype")
```

\newpage

## Group Comparison - Total Sleep

```{r sleep-comparison, fig.cap="Total sleep comparison between groups. Bars show mean ± SEM, points show individual flies."}
plotGroupComparison(monitor, 
                    metric = "total_sleep",
                    title = "Total Sleep by Genotype")
```

\newpage

## Activity Distribution

```{r activity-dist, fig.cap="Distribution of total activity across all flies."}
total_activity <- colSums(monitor@assays$mt, na.rm = TRUE)
df <- data.frame(activity = total_activity)

ggplot(df, aes(x = activity)) +
  geom_histogram(bins = 30, fill = "#2E86AB", color = "white", alpha = 0.8) +
  labs(title = "Total Activity Distribution",
       x = "Total Activity (beam crosses)",
       y = "Number of Flies") +
  theme_rmb()
```

\newpage

## Sleep Distribution

```{r sleep-dist, fig.cap="Distribution of total sleep time across all flies."}
df <- data.frame(sleep = total_sleep)

ggplot(df, aes(x = sleep)) +
  geom_histogram(bins = 30, fill = "#E94560", color = "white", alpha = 0.8) +
  labs(title = "Total Sleep Distribution",
       x = "Total Sleep (minutes)",
       y = "Number of Flies") +
  theme_rmb()
```

\newpage

## Time Series - Single Fly Activity

```{r timeseries, fig.cap="Activity time series for a single fly over the entire recording period."}
plotTimeSeries(monitor@assays$mt[, 1], 
               y_label = "Activity (beam crosses/min)",
               title = "Activity Time Series - Fly 1")
```

\newpage

# Package Functions Reference

## Core Functions

| Function | Description |
|----------|-------------|
| `createMonitor()` | Create MonitorS4 object from file |
| `processMonitor()` | Parse raw data into assays |
| `mergeMonitors()` | Combine multiple monitors |
| `removeDeadFlies()` | QC: remove inactive flies |

## Data Transforms

| Function | Description |
|----------|-------------|
| `activityToAwake()` | Convert activity to awake/sleep binary |
| `awakeToSleep()` | Flip awake to sleep binary |
| `positionFrequency()` | Count position occurrences |
| `maskByAwake()` | Position data only when awake |
| `binTimeSeries()` | Aggregate into time bins |
| `splitLightDark()` | Separate light/dark periods |

## Plotting Functions

| Function | Description |
|----------|-------------|
| `plotActivityRaster()` | Actogram-style raster plot |
| `plotSleepProfile()` | Sleep across 24h by group |
| `plotGroupComparison()` | Bar plot comparing groups |
| `plotTimeSeries()` | Line plot of time series |
| `theme_rmb()` | Publication-ready theme |
| `scale_fill_rmb()` | Custom color palette |

## Pipeline & Checkpoints

| Function | Description |
|----------|-------------|
| `runPipeline()` | Complete analysis in one command |
| `quickAnalysis()` | Fast exploratory analysis |
| `checkpoint()` | Save analysis state with plots |
| `summarizeMonitor()` | Print data summary |
| `generateReport()` | Create analysis report |

## Utilities

| Function | Description |
|----------|-------------|
| `loadMonitorFiles()` | Find monitor files |
| `loadMetadata()` | Load metadata CSVs |
| `saveResults()` | Export to RDS/CSV |
| `checkQuality()` | Data quality checks |
| `calculateZT()` | Convert to Zeitgeber time |

\newpage

# Quick Start Examples

## One-Command Pipeline

```r
# Run complete analysis
results <- runPipeline(
  data_dir = "path/to/data",
  output_dir = "path/to/output",
  experiment_name = "My Experiment",
  make_plots = TRUE,
  make_checkpoints = TRUE
)

# Access processed data
monitor <- results$monitor
```

## Quick Exploratory Analysis

```r
# Fast analysis without bells and whistles
monitor <- quickAnalysis("path/to/data")
```

## Manual Step-by-Step

```r
# Load
files <- loadMonitorFiles("path/to/data")
monitors <- lapply(files, function(f) {
  m <- createMonitor(f)
  processMonitor(m)
})

# Merge and QC
monitor <- mergeMonitors(monitors)
monitor <- removeDeadFlies(monitor)

# Sleep analysis
monitor@assays$awake <- activityToAwake(monitor@assays$mt)
monitor@assays$sleep <- awakeToSleep(monitor@assays$awake)

# Visualize
plotSleepProfile(monitor)
plotGroupComparison(monitor, metric = "total_sleep")

# Save
saveResults(monitor, "output/", prefix = "my_analysis")
```

\newpage

# Summary of Changes Made

## Files Created/Modified

1. **R/constants.R** - No more magic numbers
2. **R/MonitorClass.R** - Proper S4 class
3. **R/transforms.R** - Consolidated data transforms
4. **R/plotting.R** - Basic plots
5. **R/plotting_advanced.R** - Actograms, sleep profiles
6. **R/themes.R** - Publication themes
7. **R/checkpoints.R** - Analysis checkpoints
8. **R/pipeline.R** - One-command pipeline
9. **R/utils.R** - Helper functions

## Commits Made (on `clawd` branch)

1. Refactor: Consolidate and optimize codebase
2. Fix: plotTimeSeries day_breaks error
3. Major: Convert to proper R package structure
4. Add testthat infrastructure and tests
5. Add utility functions and build ignore
6. Add GitHub Actions CI workflows
7. Add example data and update DESCRIPTION

## Test Results

- **Monitors loaded**: 6
- **Channels before QC**: 192
- **Channels after QC**: 184 (8 dead flies removed)
- **Recording duration**: 5.8 days
- **Mean sleep**: 5649.5 minutes (68% of time)
- **Pipeline runtime**: 9.6 seconds

---

*Report generated by Catbot 🐱*
