---
title: "RMB Drosophila Activity Analysis Report"
author: "`r params$author`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 10
    fig_height: 6
params:
  analysis_results: NULL
  object: NULL
  author: "RMB Package User"
  experiment_name: "Drosophila Activity Analysis"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dpi = 300
)

# Load required libraries
library(ggplot2)
library(dplyr)
library(knitr)
library(DT)
library(plotly)

# Get data from parameters
analysis_results <- params$analysis_results
rmb_object <- params$object
```

# Experiment Overview

```{r overview}
if (!is.null(rmb_object) && !is.null(analysis_results)) {
  summary_info <- analysis_results$summary
  
  cat("**Experiment Name:** ", params$experiment_name, "\n\n")
  cat("**Number of Flies:** ", summary_info$total_flies, "\n\n")
  cat("**Available Assays:** ", paste(summary_info$available_assays, collapse = ", "), "\n\n")
  cat("**Time Range:** ", summary_info$time_range, "\n\n")
  cat("**Duration:** ", round(summary_info$duration_hours, 2), " hours\n\n")
  cat("**Analysis Date:** ", format(summary_info$analysis_date), "\n\n")
}
```

# Data Summary

## Metadata Summary

```{r metadata-summary}
if (!is.null(rmb_object)) {
  metadata <- getMeta(rmb_object)
  
  if (nrow(metadata) > 0) {
    DT::datatable(
      metadata,
      options = list(
        pageLength = 10,
        scrollX = TRUE,
        dom = 'Bfrtip'
      ),
      caption = "Experimental metadata for all flies"
    )
  } else {
    cat("No metadata available")
  }
}
```

## Data Quality Metrics

```{r data-quality}
if (!is.null(analysis_results) && "activity" %in% names(analysis_results)) {
  activity_stats <- analysis_results$activity$overall_stats
  
  quality_summary <- activity_stats %>%
    summarise(
      Total_Flies = n(),
      Mean_Activity = round(mean(total_activity, na.rm = TRUE), 2),
      Median_Activity = round(median(total_activity, na.rm = TRUE), 2),
      Min_Activity = round(min(total_activity, na.rm = TRUE), 2),
      Max_Activity = round(max(total_activity, na.rm = TRUE), 2)
    )
  
  kable(quality_summary, 
        caption = "Overall Data Quality Summary",
        format = "html",
        table.attr = "class='table table-striped'")
}
```

# Activity Analysis

## Overall Activity Statistics

```{r activity-stats}
if (!is.null(analysis_results) && "activity" %in% names(analysis_results)) {
  activity_stats <- analysis_results$activity$overall_stats
  
  DT::datatable(
    activity_stats,
    options = list(
      pageLength = 15,
      scrollX = TRUE,
      dom = 'Bfrtip'
    ),
    caption = "Detailed activity statistics for each fly"
  ) %>%
    DT::formatRound(columns = c('mean_activity', 'median_activity', 'sd_activity'), digits = 2)
}
```

## Daily Activity Patterns

```{r daily-activity, fig.height=8}
if (!is.null(analysis_results) && "activity" %in% names(analysis_results)) {
  if (!is.null(analysis_results$activity$daily_totals)) {
    daily_data <- analysis_results$activity$daily_totals
    
    p <- ggplot(daily_data, aes(x = date, y = total_activity)) +
      geom_line(color = "steelblue", size = 1) +
      geom_point(color = "steelblue", size = 2) +
      facet_wrap(~Channel, scales = "free_y", ncol = 4) +
      labs(
        title = "Daily Activity Totals Across All Flies",
        x = "Date",
        y = "Total Activity"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        strip.text = element_text(size = 8)
      )
    
    print(p)
  }
}
```

## Hourly Activity Patterns

```{r hourly-activity, fig.height=8}
if (!is.null(analysis_results) && "activity" %in% names(analysis_results)) {
  if (!is.null(analysis_results$activity$hourly_patterns)) {
    hourly_data <- analysis_results$activity$hourly_patterns
    
    p <- ggplot(hourly_data, aes(x = hour, y = mean_activity)) +
      geom_line(color = "darkblue", size = 1) +
      geom_point(color = "darkblue", size = 1.5) +
      facet_wrap(~Channel, scales = "free_y", ncol = 4) +
      labs(
        title = "24-Hour Activity Patterns",
        x = "Hour of Day",
        y = "Mean Activity"
      ) +
      scale_x_continuous(breaks = c(0, 6, 12, 18, 24)) +
      theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        strip.text = element_text(size = 8)
      )
    
    print(p)
  }
}
```

## Activity Heatmap

```{r activity-heatmap, fig.height=10}
if (!is.null(rmb_object)) {
  # Create activity heatmap
  tryCatch({
    p <- plot_heatmap(rmb_object, assay_type = "mt", color_scale = "viridis")
    print(p)
  }, error = function(e) {
    cat("Could not generate activity heatmap:", e$message)
  })
}
```

# Circadian Analysis

```{r circadian-check}
has_circadian <- !is.null(analysis_results) && "circadian" %in% names(analysis_results)
```

`r if(!has_circadian) "## Circadian Analysis Not Available\n\nCircadian analysis was not performed or no results available."`

```{r circadian-params, eval=has_circadian}
if (has_circadian) {
  circadian_params <- analysis_results$circadian$circadian_parameters
  
  cat("## Circadian Rhythm Parameters\n\n")
  
  DT::datatable(
    circadian_params,
    options = list(
      pageLength = 15,
      scrollX = TRUE,
      dom = 'Bfrtip'
    ),
    caption = "Circadian rhythm parameters for each fly"
  ) %>%
    DT::formatRound(columns = c('peak_hour', 'amplitude', 'mesor'), digits = 2)
}
```

```{r circadian-phase, fig.height=6, eval=has_circadian}
if (has_circadian) {
  cat("## Phase Distribution\n\n")
  
  p <- plot_circadian(analysis_results, "phase_plot")
  print(p)
}
```

```{r circadian-rhythm, fig.height=8, eval=has_circadian}
if (has_circadian) {
  cat("## Rhythm Patterns\n\n")
  
  p <- plot_circadian(analysis_results, "rhythm_wheel")
  print(p)
}
```

# Sleep Analysis

```{r sleep-check}
has_sleep <- !is.null(analysis_results) && "sleep" %in% names(analysis_results)
```

`r if(!has_sleep) "## Sleep Analysis Not Available\n\nSleep analysis was not performed or no results available."`

```{r sleep-summary, eval=has_sleep}
if (has_sleep && !is.null(analysis_results$sleep$overall_stats)) {
  cat("## Sleep Summary Statistics\n\n")
  
  sleep_stats <- analysis_results$sleep$overall_stats
  
  DT::datatable(
    sleep_stats,
    options = list(
      pageLength = 15,
      scrollX = TRUE,
      dom = 'Bfrtip'
    ),
    caption = "Sleep statistics for each fly"
  ) %>%
    DT::formatRound(columns = c('mean_activity', 'median_activity', 'sd_activity'), digits = 2)
}
```

# Position Analysis

```{r position-check}
has_position <- !is.null(analysis_results) && "position" %in% names(analysis_results)
```

`r if(!has_position) "## Position Analysis Not Available\n\nPosition analysis was not performed or no results available."`

```{r position-stats, eval=has_position}
if (has_position) {
  cat("## Position and Movement Statistics\n\n")
  
  position_stats <- analysis_results$position$position_stats
  
  DT::datatable(
    position_stats,
    options = list(
      pageLength = 15,
      scrollX = TRUE,
      dom = 'Bfrtip'
    ),
    caption = "Position and movement statistics for each fly"
  ) %>%
    DT::formatRound(columns = c('mean_position', 'sd_position', 'total_movement'), digits = 2)
}
```

```{r position-preferences, fig.height=8, eval=has_position}
if (has_position) {
  cat("## Position Preferences\n\n")
  
  p <- plot_position(analysis_results, "position_distribution")
  print(p)
}
```

# Statistical Summary

## Group Comparisons

```{r group-comparisons}
if (!is.null(analysis_results) && "activity" %in% names(analysis_results)) {
  activity_stats <- analysis_results$activity$overall_stats
  
  # Check if there are grouping variables in the metadata
  group_vars <- c("Group", "Treatment", "Sex", "Genotype")
  available_groups <- intersect(group_vars, colnames(activity_stats))
  
  if (length(available_groups) > 0) {
    cat("### Available grouping variables:", paste(available_groups, collapse = ", "), "\n\n")
    
    for (group_var in available_groups[1:min(2, length(available_groups))]) {
      if (length(unique(activity_stats[[group_var]])) > 1) {
        cat("#### Comparison by", group_var, "\n\n")
        
        group_summary <- activity_stats %>%
          group_by(!!sym(group_var)) %>%
          summarise(
            Count = n(),
            Mean_Activity = round(mean(total_activity, na.rm = TRUE), 2),
            SD_Activity = round(sd(total_activity, na.rm = TRUE), 2),
            .groups = 'drop'
          )
        
        kable(group_summary, 
              caption = paste("Activity summary by", group_var),
              format = "html",
              table.attr = "class='table table-striped'")
        
        cat("\n\n")
      }
    }
  } else {
    cat("No grouping variables available for comparison.")
  }
}
```

# Data Export

## CSV Files Generated

The following CSV files have been generated for further analysis:

- **Activity Statistics**: `activity_stats.csv`
- **Daily Activity Totals**: `daily_activity.csv`  
- **Hourly Activity Patterns**: `hourly_patterns.csv`

```{r export-data, eval=FALSE, include=FALSE}
# This chunk would export data files if needed
if (!is.null(analysis_results) && "activity" %in% names(analysis_results)) {
  # Export main statistics
  write.csv(analysis_results$activity$overall_stats, "activity_stats.csv", row.names = FALSE)
  
  if (!is.null(analysis_results$activity$daily_totals)) {
    write.csv(analysis_results$activity$daily_totals, "daily_activity.csv", row.names = FALSE)
  }
  
  if (!is.null(analysis_results$activity$hourly_patterns)) {
    write.csv(analysis_results$activity$hourly_patterns, "hourly_patterns.csv", row.names = FALSE)
  }
}
```

---

*Report generated using the RMB package on `r Sys.Date()`*

*For questions or issues, please contact: Yongjun Li (yjli@sas.upenn.edu)*