---
title: "Getting Started with RMB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with RMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

## Introduction

RMB (R Multi-Beam) is an R package for analyzing Drosophila activity and sleep
data from TriKinetics Multi-Beam (MB) and Single-Beam (SB) monitoring systems.

### Key Features

- **MonitorS4 Object**: Organized data structure similar to Seurat
- **Complete Pipeline**: From raw data to publication-ready figures
- **Sleep Analysis**: Standard 5-minute criterion for sleep definition
- **Checkpoints**: Save analysis state at each step
- **Beautiful Plots**: Publication-quality visualizations

## Installation

```{r, eval=FALSE}
# Install from GitHub
devtools::install_github("yjli2017/RMB")
```

## Quick Start

### Load the package

```{r, eval=FALSE}
library(RMB)
```

### Run the complete pipeline

The easiest way to analyze your data:

```{r, eval=FALSE}
# Run the full pipeline
results <- runPipeline(
  data_dir = "path/to/your/data",
  experiment_name = "My Experiment",
  make_plots = TRUE,
  make_checkpoints = TRUE
)

# Access the processed monitor
monitor <- results$monitor
```

### Or step by step

```{r, eval=FALSE}
# 1. Load data
monitor_files <- list.files("path/to/data", pattern = "Monitor.*\\.txt$", full.names = TRUE)
monitor_list <- lapply(monitor_files, function(f) {
  m <- createMonitor(f)
  m <- processMonitor(m)
  return(m)
})

# 2. Merge monitors
monitor <- mergeMonitors(monitor_list)

# 3. Quality control
monitor <- removeDeadFlies(monitor, min_movement = 10)

# 4. Sleep analysis
monitor@assays$awake <- activityToAwake(monitor@assays$mt, threshold = 5)
monitor@assays$sleep <- awakeToSleep(monitor@assays$awake)

# 5. Visualize
plotSleepProfile(monitor)
```

## Data Structure

### MonitorS4 Object

The MonitorS4 object has the following slots:

- `@assays`: List of data matrices
  - `mt`: Movement/activity data
  - `pn`: Position data (multi-beam)
  - `ct`: Cumulative data
  - `awake`: Binary awake/asleep (after sleep analysis)
  - `sleep`: Binary sleep (after sleep analysis)
- `@meta.data`: Fly metadata (phenotype, treatment, etc.)
- `@time`: Time information
- `@monitor_name`: Monitor identifier
- `@file_path`: Source file path

### Accessing Data

```{r, eval=FALSE}
# Activity data (timepoints x flies)
activity <- monitor@assays$mt

# Sleep data
sleep <- monitor@assays$sleep

# Metadata
metadata <- monitor@meta.data

# Total activity per fly
total_activity <- colSums(activity, na.rm = TRUE)

# Total sleep per fly
total_sleep <- colSums(sleep, na.rm = TRUE)
```

## Sleep Analysis

RMB uses the standard Drosophila sleep definition: **5 or more consecutive
minutes without any beam crosses**.

```{r, eval=FALSE}
# Convert activity to awake/sleep
monitor@assays$awake <- activityToAwake(monitor@assays$mt, threshold = 5)
monitor@assays$sleep <- awakeToSleep(monitor@assays$awake)

# Calculate sleep metrics
total_sleep <- colSums(monitor@assays$sleep, na.rm = TRUE)
percent_sleep <- total_sleep / nrow(monitor@assays$sleep) * 100
```

## Visualization

### Sleep Profile

```{r, eval=FALSE}
plotSleepProfile(monitor, 
                 bin_minutes = 30,
                 title = "Sleep Profile")
```

### Activity Raster

```{r, eval=FALSE}
plotActivityRaster(monitor, 
                   channel = 1,
                   title = "Activity Raster")
```

### Group Comparison

```{r, eval=FALSE}
plotGroupComparison(monitor, 
                    metric = "total_sleep",
                    title = "Total Sleep by Genotype")
```

### Summary Figure

```{r, eval=FALSE}
plotSummaryFigure(monitor, output_file = "summary.png")
```

## Checkpoints

Checkpoints save the analysis state and generate diagnostic plots:

```{r, eval=FALSE}
# Save a checkpoint
checkpoint(monitor, "after_qc", output_dir = "results/")

# This creates:
# - results/checkpoints/after_qc/summary.txt
# - results/checkpoints/after_qc/monitor.rds
# - results/checkpoints/after_qc/activity_heatmap.png
# - results/checkpoints/after_qc/activity_distribution.png
```

## Themes

RMB includes custom ggplot2 themes:

```{r, eval=FALSE}
library(ggplot2)

# Default theme
ggplot(data, aes(x, y)) + 
  geom_point() + 
  theme_rmb()

# Dark theme (for presentations)
ggplot(data, aes(x, y)) + 
  geom_point() + 
  theme_rmb_dark()

# RMB color scales
ggplot(data, aes(x, y, fill = group)) + 
  geom_bar(stat = "identity") +
  scale_fill_rmb()
```

## Exporting Data

```{r, eval=FALSE}
# Save processed monitor
saveRDS(monitor, "monitor_processed.rds")

# Export to CSV
activity_summary <- data.frame(
  fly = colnames(monitor@assays$mt),
  total_activity = colSums(monitor@assays$mt),
  total_sleep = colSums(monitor@assays$sleep)
)
write.csv(activity_summary, "activity_summary.csv", row.names = FALSE)
```

## Session Info

```{r}
sessionInfo()
```
