---
title: "Getting Started with RMB: Drosophila Activity Data Analysis"
author: "Yongjun Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with RMB: Drosophila Activity Data Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

The RMB package provides a comprehensive toolkit for analyzing Drosophila behavioral data from TriKinetics monitors. This vignette demonstrates the main functionality of the package, from loading data to generating comprehensive analysis reports.

## Key Features

- **Simple data loading**: Load and process monitor data with a single function
- **Comprehensive analysis**: Activity, sleep, circadian rhythm, and position analysis
- **Rich visualizations**: Heatmaps, time series plots, and summary charts
- **HTML reporting**: Generate professional analysis reports
- **S4 object system**: Organized data structure similar to Seurat objects

# Installation

```{r installation, eval=FALSE}
# Install development version from GitHub (when available)
devtools::install_github("username/RMB")

# Or install locally
devtools::install_local(".")
```

# Loading the Package

```{r setup}
library(RMB)
library(ggplot2)
library(dplyr)
```

# Basic Workflow

## Step 1: Load Your Data

The core function of RMB is `loadRMB()`, which takes paths to your data and metadata:

```{r load-data, eval=FALSE}
# Load data with separate data and metadata directories
data <- loadRMB(data = "./test/", metadata = "./test/metadata/")

# Or load a single monitor file with specific metadata
data <- loadRMB(data = "./test/Monitor36.txt", 
               metadata = "./test/metadata/Monitor36_metadata.csv")
```

For this vignette, we'll create some example data:

```{r create-example-data}
# This would normally be your actual data loading
# For demonstration, we'll create a minimal example structure
# In practice, you would use: data <- loadRMB(data = "your/data/path", metadata = "your/metadata/path")

# Create example metadata
example_metadata <- data.frame(
  Channel = 1:8,
  Fly = paste0("Fly_", 1:8),
  Treatment = rep(c("Control", "Treatment"), each = 4),
  Sex = rep(c("Male", "Female"), 4),
  stringsAsFactors = FALSE
)

cat("Example metadata structure:")
head(example_metadata)
```

## Step 2: Explore the Data Structure

RMB uses a custom S4 class called `MonitorS4` to organize your data:

```{r explore-structure, eval=FALSE}
# View basic information about your data
data

# Access metadata
metadata <- getMeta(data)
head(metadata)

# Check available assays
names(data@assays)

# View time information
data@time$duration_hours
```

## Step 3: Run Comprehensive Analysis

The `comprehensive_analysis()` function performs all major analyses:

```{r run-analysis, eval=FALSE}
# Run complete analysis suite
results <- comprehensive_analysis(data)

# Check what analyses were performed
names(results)
```

## Step 4: Generate Report

Create a professional HTML report with all results:

```{r generate-report, eval=FALSE}
# Generate comprehensive report
report_path <- generateReport(data, results, 
                            output_file = "my_experiment_report.html",
                            experiment_name = "Caffeine Treatment Study",
                            author = "Yongjun Li")

# Generate quick summary report
quick_report <- generateQuickReport(data, "quick_summary.html")
```

# Detailed Analysis Functions

## Activity Analysis

```{r activity-analysis, eval=FALSE}
# Detailed activity analysis
activity_results <- analyze_activity(data)

# View activity statistics
head(activity_results$overall_stats)

# Plot daily activity patterns
plot_activity_bars(activity_results, plot_type = "daily_total")

# Plot hourly patterns
plot_activity_bars(activity_results, plot_type = "hourly_pattern")
```

## Sleep Analysis

```{r sleep-analysis, eval=FALSE}
# Calculate sleep patterns
data_with_sleep <- calculate_sleep(data, sleep_threshold = 5)

# Sleep analysis
sleep_results <- analyze_activity(data_with_sleep)
```

## Circadian Analysis

```{r circadian-analysis, eval=FALSE}
# Circadian rhythm analysis
circadian_results <- analyze_circadian(data)

# View circadian parameters
head(circadian_results$circadian_parameters)

# Plot circadian patterns
plot_circadian(circadian_results, "rhythm_wheel")
plot_circadian(circadian_results, "phase_plot")
```

## Position Analysis (if position data available)

```{r position-analysis, eval=FALSE}
# Position and movement analysis
if ("pn" %in% names(data@assays)) {
  position_results <- analyze_position(data)
  
  # Plot position preferences
  plot_position(position_results, "position_distribution")
  
  # Plot movement statistics
  plot_position(position_results, "movement_stats")
}
```

# Visualization Examples

## Activity Heatmaps

```{r heatmap-example, eval=FALSE}
# Create activity heatmap
heatmap_plot <- plot_heatmap(data, assay_type = "mt")
print(heatmap_plot)

# Interactive heatmap
interactive_heatmap <- plot_heatmap(data, assay_type = "mt", interactive = TRUE)
```

## Group Comparisons

```{r group-comparisons, eval=FALSE}
# Compare groups (if grouping variables exist)
activity_results <- analyze_activity(data)

# Plot by treatment group
plot_activity_bars(activity_results, "overall_stats", group_by = "Treatment")

# Circadian analysis by group
circadian_results <- analyze_circadian(data)
plot_circadian(circadian_results, "phase_plot", group_by = "Treatment")
```

# Data Export

## Export Analysis Results

```{r data-export, eval=FALSE}
# Export all analysis data as CSV files
exported_files <- export_analysis_data(results, output_dir = "./analysis_output/")

# Files exported:
print(exported_files)
```

## Manual Data Access

```{r manual-access, eval=FALSE}
# Access specific assay data
motor_activity <- getAssay(data, "mt")
position_data <- getAssay(data, "pn")

# Access metadata
metadata <- getMeta(data)

# Export custom data
write.csv(metadata, "experiment_metadata.csv")
```

# Advanced Usage

## Working with Multiple Monitors

```{r multiple-monitors, eval=FALSE}
# Load multiple monitor files automatically
multi_data <- loadRMB(data = "./experiment_data/", 
                     metadata = "./experiment_data/metadata/")

# The function automatically detects and merges multiple monitors
```

## Custom Analysis Parameters

```{r custom-parameters, eval=FALSE}
# Custom sleep analysis
data_custom_sleep <- calculate_sleep(data, 
                                   sleep_threshold = 10,  # 10 minutes instead of 5
                                   bin_minutes = 60)      # 1-hour bins

# Custom dead fly removal
data_filtered <- loadRMB(data = "./test/", 
                        metadata = "./test/metadata/",
                        remove_dead = TRUE,
                        activity_threshold = 5,
                        consecutive_zeros = 24)
```

## Working with the MonitorS4 Object

```{r monitorS4-usage, eval=FALSE}
# Create custom MonitorS4 object
custom_object <- MonitorS4(
  meta.data = my_metadata,
  assays = list(mt = my_motor_data),
  active.assay = "mt"
)

# Add new assay data
custom_object <- setAssay(custom_object, "custom_metric", my_custom_data)

# Modify metadata
new_metadata <- getMeta(custom_object)
new_metadata$new_column <- "new_value"
custom_object <- setMeta(custom_object, new_metadata)
```

# Tips and Best Practices

## File Organization

Organize your data files as follows:

```
experiment_directory/
├── Monitor36.txt
├── Monitor37.txt
├── Monitor38.txt
└── metadata/
    ├── Monitor36_metadata.csv
    ├── Monitor37_metadata.csv
    └── Monitor38_metadata.csv
```

## Metadata Requirements

Your metadata files should include:

- **Channel**: Channel number (1-32 for most monitors)
- **Treatment**: Experimental condition
- **Sex**: Male/Female
- **Genotype**: Genetic background
- **Date**: Experiment date
- **Other relevant experimental variables**

## Quality Control

```{r quality-control, eval=FALSE}
# Check data quality before analysis
cat("Total flies:", nrow(getMeta(data)))
cat("Duration:", data@time$duration_hours, "hours")
cat("Available assays:", names(data@assays))

# Remove dead flies with custom thresholds
data_qc <- remove_dead_flies(data, activity_threshold = 2, consecutive_zeros = 18)
```

# Troubleshooting

## Common Issues

1. **Missing metadata files**: Use `auto_generate_metadata = TRUE` in `loadRMB()`
2. **Package dependencies**: Run the included `install_deps.R` script
3. **Memory issues with large datasets**: Process monitors individually then merge

## Getting Help

```{r help, eval=FALSE}
# Function documentation
?loadRMB
?comprehensive_analysis
?generateReport

# Package help
help(package = "RMB")
```

# Conclusion

The RMB package provides a streamlined workflow for Drosophila behavioral data analysis. The two-step process of `loadRMB()` followed by `generateReport()` can handle most analysis needs, while the individual analysis functions provide flexibility for custom workflows.

For more advanced usage and specific analysis questions, please refer to the function documentation or contact Yongjun Li (yjli@sas.upenn.edu).

---

*This vignette was generated using RMB version `r packageVersion("RMB")`*